Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-05-25 nemolite2d_perf.tex
nemolite2d-paper: updates to abstract and conclusions

We present an approach which we call \psykal\ that aims to achieve
portable performance for parallel, finite-difference Ocean models.  In
\psykal\ the code related to the underlying science is formally
separated from code related to parallelisation and single-core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
and for optimisation specialists to be able to tailor the code for a
particular machine independently of the science code. We have taken
the free-surface part of the NEMO ocean model and created a new,
shallow-water model named NEMOLite2D. In doing this we have a code
which is of a manageable size and yet which incorporates elements of
full ocean models (input/output, boundary conditions, \textit{etc.}).
We have then manually constructed a \psykal\ version of this code and
investigated the transformations that must be applied to the
middle/PSy layer in order to achieve good performance, both serial and
parallel. We have produced versions of the PSy layer parallelised with
both OpenMP and OpenACC; in both cases we were able to leave the
natural-science parts of the code unchanged while achieving good
performance on both multi-core CPUs and GPUs. In quantifying whether
or not the obtained performance is `good' we also consider the
limitations of the basic roofline model and attempt to improve on it
by generating kernel-specific CPU ceilings.

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Performance, Code-generation, Finite-difference, NEMO, OpenMP, OpenACC
\end{IEEEkeywords}

@@

In a previous paper~\cite{shallow_psykal} we introduced a possible
approach to tackling this problem which we term \psykal\ (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to the `Shallow' shallow-water model. We found
that although the re-structuring did initially incur a sizeable
performance penalty, it was possible to transform the resulting code
to recover performance (for a variety of CPU/compiler combinations)
while obeying the \psykal\ separation of concerns. In this work we
move to looking at portable parallel performance within the
\psykal\ approach.

@@

For this work we have used a program, `NEMOLite2D,' developed by
ourselves.  NEMOLite2D is a vertically-averaged version of NEMO
(Nucleus for European Modelling of the Ocean~\cite{nemo}), retaining
only its dynamical part. The whole model system is represented through
one continuity equation \eqref{continuity} (for the update of the
sea-surface height) and two vertically-integrated momentum equations
\eqref{momentum} (for the two velocity components, respectively).

@@

The advantage of the \psykal\ re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to port the Fortran
NEMOLite2D to GPU we used OpenACC directives as opposed to the OpenMP
directives used for the multi-core CPU. We also experimented with
using CUDA directly in order to quantify any performance penalty
incurred.

The introduction of these directives proved to be simpler than
producing the OpenMP version of NEMOLite2D {\bf - Jeremy to clarify
  whether this is true}.  Consequently we only show the performance of
{\bf xxx} versions of the code in
Figure~\ref{FIG_gpu_opt_stages}. {\bf discuss performance drop in
  OpenACC version c.f. CUDA.}

@@

Once the problem size is increased to $128^2$, a single time-step
takes roughly 200~$\mu$s and only the Intel-compiled OpenMP version is
comparable in performance to the OpenACC version. For all of the
larger problem sizes plotted in Figure~\ref{FIG_cpu_cf_gpu} the GPU
version is considerably faster than the CPU.  For problem sizes of
$1024^2$ and greater, the 30MB cache of the Ivy Bridge CPU is
exhausted and performance becomes limited by the bandwidth to main
memory. At this stage the OpenACC version of the code on the GPU is
some {\bf 3.7 (this is the factor for the CUDA version!)} times faster
than the OpenMP version on the CPU. {\bf Can we relate this to the
  difference in peak memory b/w between the GPU and CPU?}

@@

In Figure~\ref{FIG_roofline} we show the performance of kernels from
both Shallow and NEMOLite2D on the roofline model for the E5-1620 CPU.
This demonstrates that the Shallow kernel is achieving a performance
roughly consistent with saturating the available memory bandwidth. In
contrast, the kernel taken from NEMOLite2D is struggling to reach a
performance consistent with saturating the bandwidth to main
memory. We experimented with reducing the problem size (so as to
ensure it fitted within cache) but that did not significantly improve
the performance of the kernel. This then points to more fundamental issues
with the way that the kernel is implemented which are not captured in the
simple roofline model.

@@

We have used a new and un-optimised, two-dimension model extracted
from the NEMO ocean model for this work. As a consequence of this, the
introduction of the \psykal\ separation of concerns followed by
suitable transformations of the PSy Layer is actually found to improve
performance. This is in contrast to our previous
experience~\cite{shallow_psykal} with tackling the Shallow code which
has had been optimised over the course of several years. In that case
we were able to recover more than 90\% of the performance of the
original for the Gnu, Intel, Cray and IBM compilers on both Intel and
IBM Power CPUs.

Investigation of the absolute serial performance of the NEMLite2D code
using the roofline model revealed that it was still significantly
below any of the traditional roofline ceilings. We have developed a
code-analysis tool~\cite{habakkuk} that is capable of providing more
accurate ceilings by analysing the nature of the floating point
computations performed by a kernel. In future work we aim to extend
this tool such that it can be applied to code parallelised using
OpenMP.

The application of code transformations to the middle/PSy layer is key
to the performance of the \psykal\ version of a code. For both
NEMLite2D and Shallow we have found that for serial performance, the
most important transformation is that of in-lining the kernel source
at the call site. However, for the more complex NEMOLite2D code the
Cray compiler also had to be coerced into performing SIMD
vectorisation through the use of souce-code directives.

In this work we have also demonstrated the introduction of parallelism
into the PSy layer with both OpenMP and OpenACC directives. In both
cases we were able to leave the natural-science parts of the code
unchanged. For OpenMP we achieved a parallel efficiency of $\sim70\%$
on 12 threads by enclosing the body of the (single) PSy layer routine
within a single PARALLEL region. Removal of unecessary synchronisation
points through use of the NOWAIT clause boosted 12-thread performance
by approximately 10\% with the Gnu and Cray compilers. The
\psykal\ re-structuring of this (admittedly small) code was not found
to pose any problems for the introduction of performant OpenMP.

{\bf Some words about how good the OpenACC, \psykal\ version is compared
to the CUDA version.}

Having demonstrated that the \psykal\ separation of concerns may be
applied to 2D finite-difference codes with minimal performance loss,
our next steps will be; 1) to consider the automatic generation of the
PSy layer and 2) to look at extending the approach to the full NEMO
model (i.e. three dimensions). In future work we will analyse the
performance of a domain-specific compiler that performs the automatic
generation of the PSy layer. This compiler (which we have named
`PSyclone') is currently under development.

>>>>

We present an approach which we call \psykal\ that is designed to achieve
portable performance for parallel, finite-difference Ocean models.  In
\psykal\ the code related to the underlying science is formally
separated from code related to parallelisation and single-core
optimisations. This separation of concerns allows scientists to code
their science independently of the underlying hardware architecture
and for optimisation specialists to be able to tailor the code for a
particular machine independently of the science code. We have taken
the free-surface part of the NEMO ocean model and created a new,
shallow-water model named NEMOLite2D. In doing this we have a code
which is of a manageable size and yet which incorporates elements of
full ocean models (input/output, boundary conditions, \textit{etc.}).
We have then manually constructed a \psykal\ version of this code and
investigated the transformations that must be applied to the
middle/PSy layer in order to achieve good performance, both serial and
parallel. We have produced versions of the PSy layer parallelised with
both OpenMP and OpenACC; in both cases we were able to leave the
natural-science parts of the code unchanged while achieving good
performance on both multi-core CPUs and GPUs. In quantifying whether
or not the obtained performance is `good' we also consider the
limitations of the basic roofline model and improve on it
by generating kernel-specific CPU ceilings.

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Performance, Code-generation, Finite-difference, NEMO, OpenMP, OpenACC, roofline
\end{IEEEkeywords}

@@

In a previous paper~\cite{shallow_psykal} we introduced a possible
approach to tackling this problem which we term \psykal\ (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to the `Shallow' shallow-water model
(https://puma.nerc.ac.uk/trac/GOcean). We found that although the
re-structuring did initially incur a sizeable performance penalty, it
was possible to transform the resulting code to recover performance
(for a variety of CPU/compiler combinations) while obeying the
\psykal\ separation of concerns. In this work we move to looking at
portable parallel performance within the \psykal\ approach.

@@

For this work we have used a program, `NEMOLite2D,' developed by
ourselves (https://puma.nerc.ac.uk/trac/GOcean).  NEMOLite2D is a
vertically-averaged version of NEMO (Nucleus for European Modelling of
the Ocean~\cite{nemo}), retaining only its dynamical part. The whole
model system is represented through one continuity equation
\eqref{continuity} (for the update of the sea-surface height) and two
vertically-integrated momentum equations \eqref{momentum} (for the two
velocity components, respectively).

@@

The advantage of the \psykal\ re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the \psykal\ re-structuring, we
had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the \psykal
apprach. He also experimented with using CUDA directly within the
original form of NEMOLite2D in order to quantify any performance
penalty incurred by taking the \psykal/OpenACC approach.

The performance of these two versions of the code is plotted in
Figure~\ref{FIG_gpu_opt_stages}.  Since the \psykal restructuring of
NEMOLite2D did not prove to be a barrier to achieving a performant
code when using OpenACC directives, we only report the performance of
the \psykal\ OpenACC version. For problem sizes up to $256^2$ the
difference in the performance of the two versions is less than 5\% and
for the larger domains it is at most 11\%. This demonstrates that the
performance cost of the \psykal\ approach in utilising a GPU for
NEMOLite2D is minimal.

@@

Once the problem size is increased to $128^2$, a single time-step
takes roughly 200~$\mu$s and only the Intel-compiled OpenMP version is
comparable in performance to the OpenACC version. For all of the
larger problem sizes plotted in Figure~\ref{FIG_cpu_cf_gpu} the GPU
version is considerably faster than the CPU.  For problem sizes of
$1024^2$ and greater, the 30MB cache of the Ivy Bridge CPU is
exhausted and performance becomes limited by the bandwidth to main
memory. At this stage the OpenACC version of the code on the GPU is
some 3.4 times faster than the best OpenMP version on the CPU.

%Peak memory bandwidth on the K40 GPU is 288 GB/s and we measured
%(using STREAM2) bandwidths of 46.0 and 20.5 GB/s to L3 and main
%memory, respectively on the CPU.

@@

In Figure~\ref{FIG_roofline} we show the performance of kernels from
both Shallow and NEMOLite2D on the roofline model for an Intel E5-1620 CPU.
This demonstrates that the Shallow kernel is achieving a performance
roughly consistent with saturating the available memory bandwidth. In
contrast, the kernel taken from NEMOLite2D is struggling to reach a
performance consistent with saturating the bandwidth to main
memory. We experimented with reducing the problem size (so as to
ensure it fitted within cache) but that did not significantly improve
the performance of the kernel. This then points to more fundamental issues
with the way that the kernel is implemented which are not captured in the
simple roofline model.

@@

We have used a new and un-optimised, two-dimensional model extracted
from the NEMO ocean model for this work. As a consequence of this, the
introduction of the \psykal\ separation of concerns followed by
suitable transformations of the PSy Layer is actually found to improve
performance. This is in contrast to our previous
experience~\cite{shallow_psykal} with tackling the Shallow code which
has been optimised over many years. In that case we were able to
recover (to within a few percent) the performance of the original and
in some cases exceed it, in spite of limiting ourselves to
transformations which replicated the structure of the original,
optimised code.

Investigation of the absolute serial performance of the NEMOLite2D
code using the roofline model revealed that it was still significantly
below any of the traditional roofline ceilings. We have developed
Habakkuk (https://github.com/arporter/habakkuk), a code-analysis tool
that is capable of providing more realistic ceilings by analysing the
nature of the floating point computations performed by a kernel. The
bounds produced by Habakkuk are in good agreement with the measured
performance of the principal (Momentum) kernel in NEMOLite2D. In
future work we aim to extend this tool to account for SIMD operations
and make it applicable to code parallelised using OpenMP.

The application of code transformations to the middle/PSy layer is key
to the performance of the \psykal\ version of a code. For both
NEMOLite2D and Shallow we have found that for serial performance, the
most important transformation is that of in-lining the kernel source
at the call site, i.e. within the PSy layer. (Although we have done
this in-lining manually for this work, our aim is that in future, such
transformations will be performed automatically at compile-time and
therefore do not affect the code that a scientist writes.)
For the more complex NEMOLite2D code the Cray compiler also
had to be coerced into performing SIMD vectorisation through the use
of souce-code directives.

In this work we have also demonstrated the introduction of parallelism
into the PSy layer with both OpenMP and OpenACC directives. In both
cases we were able to leave the natural-science parts of the code
unchanged. For OpenMP we achieved a parallel efficiency of $\sim70\%$
on 12 threads by enclosing the body of the (single) PSy layer routine
within a single PARALLEL region. Removal of unecessary synchronisation
points through use of the NOWAIT clause boosted 12-thread performance
by approximately 10\% with the Gnu and Cray compilers. The
\psykal\ re-structuring of this (admittedly small) code was not found
to pose any problems for the introduction of performant OpenMP.
Similarly, we have also demonstrated good GPU performance using OpenACC
in a \psykal\ version of the code.

This paper demonstrates that the \psykal\ separation of concerns may
be applied to 2D finite-difference codes without loss of
performance. We have also shown that the resulting code is amenable to
efficient parallelisation on both GPU and shared-memory CPU systems.
This then means that it is possible to achieve performance portability
while maintaining single-source science code.

Our next steps will be first, to consider the automatic generation of the
PSy layer and second, to look at extending the approach to the full NEMO
model (i.e. three dimensions). In future work we will analyse the
performance of a domain-specific compiler that performs the automatic
generation of the PSy layer. This compiler (which we have named
`PSyclone', see https://github.com/stfc/psyclone) is currently under
development.
